The Magnetorotational Instability Prefers Three Dimensions

### 2019-11-10 paper/paper.tex
just kidding. This is the last set of changes

\renewcommand{\l}{\ell}
\renewcommand{\ell}{\ensuremath{k_{y}}}
\newcommand{\txt}[1]{\quad \text{#1} \quad }

@@

\subsection{Asymptotic Calculation}
\label{sec:asymp}
The MRI's preference for three-dimensional modes can be predicted analytically.
We analyse equations~(\ref{eq:mhd}-\ref{eq:divu}) asymptotically close to $S_{c}$, and compute the leading-order correction to the growth rate from three-dimensional effects assuming $\SSC=1+\epsilon^{2}R$, $k_{z}\sim\mathcal{O}(\epsilon)$, $\sigma\sim{k_{y}}\sim\mathcal{O}(\epsilon^{2})$. 
Here, we outline the asymptotic calculation for the leading order correction to the 2D growth rates when accounting for three dimensional effects via finite $k_{y}$.
Being in the rotating frame means the shear has no net
\Beq
\int_{-d/2}^{d/2} V_{0,y}(x) \text{d} x  \ = \ 0 \quad \implies  \quad V_{0,y}(x) \ = \ S\, x
\Eeq
The full linear ideal equations are:
\Beq\label{u-eq}
\pd{t} v_{x} + S x \pd{y} v_{x} - f v_{y} + \pd{x} p - B_{0} \pd{z} b_{x} &=& 0\\
\pd{t} v_{y} + S x \pd{y} v_{y} + (f+S) v_{x} + \pd{y} p - B_{0} \pd{z} b_{y} &=& 0\label{v-eq} \\
\pd{t} v_{z} + S x \pd{y} v_{z} + \pd{z} p - B_{0} \pd{z} b_{z} &=& 0 \\
\pd{x} v_{x} + \pd{y} v_{y} + \pd{z} v_{z}  &=& 0\\
\pd{t} b_{x} + S x \pd{y} b_{x} - B_{0} \pd{z} v_{x} &=& 0\\
\pd{t} b_{y} + S x \pd{y} b_{y} - B_{0} \pd{z} v_{y} - S b_{x}  &=& 0\\
\pd{t} b_{z} + S x \pd{y} b_{z} - B_{0} \pd{z} v_{z} &=& 0 \label{bz-eq}
\Eeq
The divergence-free condition for the velocity implies the same for the magnetic field.
The system is 2nd order in $x$ because $\pd{x}$ only appears in two places.
All variables take the form
\Beq
p = \hat{P}(x) e^{ i ( \omega t + k_{z} z + \ell y ) },
\Eeq
where we note that for this section, we use the complex frequency $\omega = -i \sigma$ for the time dependence.

To make analytical progress, define the frequency ``parameters'' 
\Beq
\omega_{S}(x) \ \equiv \ \omega + S x \ell, \quad \omega_{A}  \ \equiv \ B_{0} k.
\Eeq
We can use equations (\ref{v-eq}--\ref{bz-eq}) to find all amplitudes in terms of $\hat{V}_{x}(x)$ and $\hat{V}_{x}'(x)$. 
\Beq
\hat{V}_{y}(x) &=& \frac{i \left(\ell 
   \hat{V}_{x}'(x) + k_{z}^2 \frac{ \left((f+S) \omega
   _S^2-S \omega _A^2\right)}{\omega_{S}
   \left(\omega_{S}^2-\omega _A^2\right)} \hat{V}_{x}(x)\right)}{k_{z}^2+\ell ^2} \\
\hat{V}_{z}(x) &=& \frac{i k_{z} \left(\hat{V}_{x}'(x)- \frac{ \left((f+S) \omega
   _S^2-S \omega _A^2\right)}{\omega_{S}
   \left(\omega_{S}^2-\omega _A^2\right)}  \ell 
   \hat{V}_{x}(x)\right)}{k_{z}^2+\ell ^2} \\
   \hat{P}(x) &=& \frac{i \left(\omega _A^2-\omega_{S}^2\right)
   \left(\hat{V}_{x}'(x) -  \ell \frac{ \left((f+S) \omega
   _S^2-S \omega _A^2\right)}{\omega_{S}
   \left(\omega_{S}^2-\omega _A^2\right)}
   \hat{V}_{x}(x)\right)}{\left(k_{z}^2+\ell ^2\right)
   \omega_{S}} \\
   \hat{B}_{x}(x) &=& \frac{\omega _A \hat{V}_{x}(x)}{\omega_{S}} \\
   \hat{B}_{y}(x) &=&\frac{\omega _A \hat{V}_{y}(x)}{\omega_{S}} -\frac{i S \omega _A \hat{V}_{x}(x)}{\omega_{S}^2} \\
   \hat{B}_{z}(x) &=& \frac{\omega _A \hat{V}_{z}(x)}{\omega_{S}}
   \Eeq
One can substitute everything into equation (\ref{u-eq}) to get a second-order equation for $\hat{V}_{x}(x)$ of the general form 
\Beq
 \hat{V}_{x}''(x) + 2 C_{1}(x) \hat{V}_{x}'(x) + C_{0}(x) \hat{V}_{x}(x) \ = \ 0,
\Eeq
where
\Beq
   C_{1}(x) &=& \frac{ S \ell  \omega _A^2}{\omega_{S} ( \omega_{S}^2-\omega _A^2 )} \\
   C_{0}(x) &=&  \frac{\frac{f^{2} k_{z}^2 \omega_{S}^{2}}{\omega_{S}^{2}-\omega
   _A^2}+f  S  k_{z}^2 -\frac{2 S^2  k_{y}^2 \omega _A^2}{\omega_{S}^{2}}}{\omega_{S}^2-\omega _A^2}  -\left(k_{z}^2+\ell^2 \right) .
\Eeq

We can eliminate the first-order term via
\Beq
\hat{V}_{x}(x) \ = \ \chi(x)\, \psi(x) \quad \text{where} \quad \frac{\chi'(x)}{\chi(x)} \  = \ - C_{1} (x)
\Eeq
\Beq
-\psi''(x) + \Phi(x) \psi(x) \ = \ 0 \label{Schr-eq}
\Eeq
\Beq
\Phi(x)  \ = \ C_{1}'(x) + C_{1} (x)^{2} - C_{0}(x).
\Eeq
Finally, by putting everything together 
\Beq
\Phi(x) \ = \ \frac{S  \left(f k_{z}^2-S \ell
   ^2\right) \omega _A^2 -f k_{z}^2 (f+S) \omega
   _S^2}{\left(\omega _A^2-\omega
   _S^2\right)^2} + k_{z}^2 + \ell^{2}.
\Eeq
The denominator of $\Phi(x)$ cannot vanish if $\omega_{S}=-i \gamma+Sx\ell$, with $Re(\gamma)\neq0$. 
We want to solve the Schr\"{o}dinger-type equation~(\ref{Schr-eq}) with the boundary conditions 
\Beq
\psi(x=\pm d/2) = 0.
\Eeq
The boundary conditions for $\psi(x)$ are the same as $V_{x}(x)$ because 
\Beq
\chi(x) \ \propto \ \frac{\omega_{S}(x)}{\sqrt{\omega_{S}(x)^2-\omega _A^2}} \  \ne \ 0 \quad \text{for} \quad \gamma \ \ne \ 0.
\Eeq
We solve perturbatively near the critical values for the 2D instability. We rescale each parameter in terms of a bookkeeping parameter, $\eps$.
\Beq
S  \ \to \  - \frac{\pi ^2 B_{0}^2}{d^2 f} ( 1 + \eps^{2} R), \quad \omega \ \to \ \eps^{2} \omega , \quad k_{z} \ \to \ \eps k_{z}, \quad \ell \ \to \ \eps^{2} \ell,
\Eeq 
keeping in mind
\Beq
\Ro  \ \equiv  \ \frac{\pi ^2 B_{0}^2}{d^2 f^{2}}.
\Eeq
Expanding equation~(\ref{Schr-eq}) to $\mathcal{O}(\eps^{2})$, we arrive at
\Beq
\Phi(x) \ = \ \frac{\pi^{2}}{d^{2}} + \eps^{2} \frac{B_{0}^2 f^2 \Ro\, \left(k_{z}^2 R+\Ro\, \ell ^2\right) - B_{0}^4 k^4 +f^2 (\Ro+1) (\omega -f\, \Ro\, x\, \ell
   )^2  }{B_{0}^4 k_{z}^2}.
\Eeq
The leading order balance is
\Beq
\psi_{0}''(x) + \frac{\pi^{2}}{d^{2}}  \psi_{0}(x) \ = \ 0  \quad \implies \quad \psi_{0}(x) \ = \ \cos(\pi x / d),
\Eeq
and next order is
\Beq
\psi_{2}''(x) + \frac{\pi^{2}}{d^{2}}  \psi_{2}(x) + \Phi_{2}(x)\psi_{0}(x)\ = \ 0.
\Eeq
The solvability condition for $\psi_{2}(x)$ is 
\Beq
\int_{-d/2}^{d/2} \Phi_{2}(x)\psi_{0}(x)^{2} \text{d} x \ = \ 0.
\Eeq
This yields our final result
\begin{equation}\label{eq:asymp}
\sigma^{2}\ = \ - \omega^{2} \ = \  \frac{\Ro\,B_{0}^{2}}{(\Ro+1)}\left[ k_{z}^{2}\left(R-\frac{d^2}{\pi^{2}}k_{z}^{2}\right)+\Upsilon\,k_{y}^{2}\right]+\ldots,
\end{equation}
where
\begin{equation}
\Upsilon \ = \  \frac{\left(\pi^{2}+6\right) \Ro+\pi^{2}-6}{12} \ \approx \ 1.31\quad\text{for}\quad{\Ro} \ = \ 0.75,
\end{equation}
and we have switched back to $\sigma$ for time dependence.
The first term in equation~(\ref{eq:asymp}) results from the two-dimensional calculation.
The second term is positive definite: it \emph{always} leads to enhanced growth rates when $k_{y}\neq0$ and ultraviolet divergence in the absence of higher-order effects.

>>>>

\newcommand{\txt}[1]{\quad \text{#1} \quad }

@@

\subsection{Asymptotic Calculation}
\label{sec:asymp}
The MRI's preference for three-dimensional modes can be predicted analytically.
We analyse equations~(\ref{eq:mhd}-\ref{eq:divu}) asymptotically close to $S_{c}$, and compute the leading-order correction to the growth rate from three-dimensional effects assuming $\SSC=1+\epsilon^{2}R$, $k_{z}\sim\mathcal{O}(\epsilon)$, $\sigma\sim{k_{y}}\sim\mathcal{O}(\epsilon^{2})$. 
Here, we outline the asymptotic calculation for the leading order correction to the 2D growth rates when accounting for three dimensional effects via finite $k_{y}$.
Being in the rotating frame means the shear has no net
\Beq
\int_{-d/2}^{d/2} V_{0,y}(x) \text{d} x  \ = \ 0 \quad \implies  \quad V_{0,y}(x) \ = \ S\, x
\Eeq
The full linear ideal equations are:
\Beq\label{u-eq}
\pd{t} v_{x} + S x \pd{y} v_{x} - f v_{y} + \pd{x} p - B_{0} \pd{z} b_{x} &=& 0\\
\pd{t} v_{y} + S x \pd{y} v_{y} + (f+S) v_{x} + \pd{y} p - B_{0} \pd{z} b_{y} &=& 0\label{v-eq} \\
\pd{t} v_{z} + S x \pd{y} v_{z} + \pd{z} p - B_{0} \pd{z} b_{z} &=& 0 \\
\pd{x} v_{x} + \pd{y} v_{y} + \pd{z} v_{z}  &=& 0\\
\pd{t} b_{x} + S x \pd{y} b_{x} - B_{0} \pd{z} v_{x} &=& 0\\
\pd{t} b_{y} + S x \pd{y} b_{y} - B_{0} \pd{z} v_{y} - S b_{x}  &=& 0\\
\pd{t} b_{z} + S x \pd{y} b_{z} - B_{0} \pd{z} v_{z} &=& 0 \label{bz-eq}
\Eeq
The divergence-free condition for the velocity implies the same for the magnetic field.
The system is 2nd order in $x$ because $\pd{x}$ only appears in two places.
All variables take the form
\Beq
p = \hat{P}(x) e^{ i ( \omega t + k_{z} z + k_{y} y ) },
\Eeq
where we note that for this section, we use the complex frequency $\omega = -i \sigma$ for the time dependence.

To make analytical progress, define the frequency ``parameters'' 
\Beq
\omega_{S}(x) \ \equiv \ \omega + S x k_{y}, \quad \omega_{A}  \ \equiv \ B_{0} k.
\Eeq
We can use equations (\ref{v-eq}--\ref{bz-eq}) to find all amplitudes in terms of $\hat{V}_{x}(x)$ and $\hat{V}_{x}'(x)$. 
\Beq
\hat{V}_{y}(x) &=& \frac{i \left(k_{y} 
   \hat{V}_{x}'(x) + k_{z}^2 \frac{ \left((f+S) \omega
   _S^2-S \omega _A^2\right)}{\omega_{S}
   \left(\omega_{S}^2-\omega _A^2\right)} \hat{V}_{x}(x)\right)}{k_{z}^2+k_{y} ^2} \\
\hat{V}_{z}(x) &=& \frac{i k_{z} \left(\hat{V}_{x}'(x)- \frac{ \left((f+S) \omega
   _S^2-S \omega _A^2\right)}{\omega_{S}
   \left(\omega_{S}^2-\omega _A^2\right)}  k_{y} 
   \hat{V}_{x}(x)\right)}{k_{z}^2+k_{y} ^2} \\
   \hat{P}(x) &=& \frac{i \left(\omega _A^2-\omega_{S}^2\right)
   \left(\hat{V}_{x}'(x) -  k_{y} \frac{ \left((f+S) \omega
   _S^2-S \omega _A^2\right)}{\omega_{S}
   \left(\omega_{S}^2-\omega _A^2\right)}
   \hat{V}_{x}(x)\right)}{\left(k_{z}^2+k_{y} ^2\right)
   \omega_{S}} \\
   \hat{B}_{x}(x) &=& \frac{\omega _A \hat{V}_{x}(x)}{\omega_{S}} \\
   \hat{B}_{y}(x) &=&\frac{\omega _A \hat{V}_{y}(x)}{\omega_{S}} -\frac{i S \omega _A \hat{V}_{x}(x)}{\omega_{S}^2} \\
   \hat{B}_{z}(x) &=& \frac{\omega _A \hat{V}_{z}(x)}{\omega_{S}}
   \Eeq
One can substitute everything into equation (\ref{u-eq}) to get a second-order equation for $\hat{V}_{x}(x)$ of the general form 
\Beq
 \hat{V}_{x}''(x) + 2 C_{1}(x) \hat{V}_{x}'(x) + C_{0}(x) \hat{V}_{x}(x) \ = \ 0,
\Eeq
where
\Beq
   C_{1}(x) \ = \  \frac{ S k_{y}  \omega _A^2}{\omega_{S} ( \omega_{S}^2-\omega _A^2 )} , \quad C_{0}(x) \ = \   \frac{\frac{f^{2} k_{z}^2 \omega_{S}^{2}}{\omega_{S}^{2}-\omega
   _A^2}+f  S  k_{z}^2 -\frac{2 S^2  k_{y}^2 \omega _A^2}{\omega_{S}^{2}}}{\omega_{S}^2-\omega _A^2}  -\left(k_{y}^{2} + k_{z}^2 \right) .
\Eeq
We can eliminate the first-order term via
\Beq
\hat{V}_{x}(x) \ = \ \chi(x)\, \psi(x) \quad \text{where} \quad \frac{\chi'(x)}{\chi(x)} \  = \ - C_{1} (x).
\Eeq
Therefore 
\Beq
-\psi''(x) + \Phi(x) \psi(x) \ = \ 0 \label{Schr-eq}, \quad \text{where}\quad
\Phi(x)  \ = \ C_{1}'(x) + C_{1} (x)^{2} - C_{0}(x).
\Eeq
Finally, by putting everything together 
\Beq
\Phi(x) \ = \ \frac{S  \left(f k_{z}^2-S k_{y}
   ^2\right) \omega _A^2 -f k_{z}^2 (f+S) \omega
   _S^2}{\left(\omega _A^2-\omega
   _S^2\right)^2} + k_{y}^{2} + k_{z}^2 .
\Eeq
The denominator of $\Phi(x)$ cannot vanish if $\omega_{S}=-i \gamma+Sxk_{y}$, with $Re(\gamma)\neq0$. 
We want to solve the Schr\"{o}dinger-type equation~(\ref{Schr-eq}) with the boundary conditions 
\Beq
\psi(x=\pm d/2) = 0.
\Eeq
The boundary conditions for $\psi(x)$ are the same as $V_{x}(x)$ because 
\Beq
\chi(x) \ \propto \ \frac{\omega_{S}(x)}{\sqrt{\omega_{S}(x)^2-\omega _A^2}} \  \ne \ 0 \quad \text{for} \quad \gamma \ \ne \ 0.
\Eeq
We solve perturbatively near the critical values for the 2D instability. We rescale each parameter in terms of a bookkeeping parameter, $\eps$.
\Beq
S  \ \to \  - \frac{\pi ^2 B_{0}^2}{d^2 f} ( 1 + \eps^{2} R), \quad \omega \ \to \ \eps^{2} \omega , \quad k_{z} \ \to \ \eps k_{z}, \quad k_{y} \ \to \ \eps^{2} k_{y},
\Eeq 
keeping in mind
\Beq
\Ro  \ \equiv  \ \frac{\pi ^2 B_{0}^2}{d^2 f^{2}}.
\Eeq
Expanding equation~(\ref{Schr-eq}) to $\mathcal{O}(\eps^{2})$, we arrive at
\Beq
\Phi(x) \ = \ \frac{\pi^{2}}{d^{2}} + \eps^{2} \frac{B_{0}^2 f^2 \Ro\, \left(k_{z}^2 R+\Ro\, k_{y} ^2\right) - B_{0}^4 k^4 +f^2 (\Ro+1) (\omega -f\, \Ro\, x\, k_{y}
   )^2  }{B_{0}^4 k_{z}^2}.
\Eeq
The leading order balance is
\Beq
\psi_{0}''(x) + \frac{\pi^{2}}{d^{2}}  \psi_{0}(x) \ = \ 0  \quad \implies \quad \psi_{0}(x) \ = \ \cos(\pi x / d),
\Eeq
and next order is
\Beq
\psi_{2}''(x) + \frac{\pi^{2}}{d^{2}}  \psi_{2}(x) + \Phi_{2}(x)\psi_{0}(x)\ = \ 0.
\Eeq
The solvability condition for $\psi_{2}(x)$ is 
\Beq
\int_{-d/2}^{d/2} \Phi_{2}(x)\psi_{0}(x)^{2} \text{d} x \ = \ 0.
\Eeq
This yields our final result
\begin{equation}\label{eq:asymp}
\sigma^{2}\ = \ - \omega^{2} \ = \  \frac{\Ro\,B_{0}^{2}}{(\Ro+1)}\left[ k_{z}^{2}\left(R-\frac{d^2}{\pi^{2}}k_{z}^{2}\right)+\Upsilon\,k_{y}^{2}\right]+\ldots,
\end{equation}
where
\begin{equation}
\Upsilon \ = \  \frac{\left(\pi^{2}+6\right) \Ro+\pi^{2}-6}{12} \ \approx \ 1.31\quad\text{for}\quad{\Ro} \ = \ 0.75,
\end{equation}
and we have switched back to $\sigma$ for time dependence.
The first term in equation~(\ref{eq:asymp}) results from the two-dimensional calculation.
The second term is positive definite: it \emph{always} leads to enhanced growth rates when $k_{y}\neq0$ and ultraviolet divergence in the absence of higher-order effects.
